---
title: "Stat 706 Final Project: Investigating Predictors of Movie Profitability"
subtitle: Nickhil Sethi
output:
  pdf_document: default
  word_document: default
  
# TODO
# - Prediction intervals
# - Look at outliers and try to find a pattern
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(printr)
library(faraway)

# global variable, don't want to echo this
GENRE_COLUMNS <- c( "genre_thriller", "genre_war", "genre_fantasy", "genre_documentary", "genre_gohands", "genre_action", "genre_family", "genre_crime", "genre_comedy", "genre_foreign", "genre_adventure", "genre_mystery", "genre_science_fiction", "genre_horror", "genre_drama", "genre_animation", "genre_romance", "genre_western", "genre_history", "genre_tv_movie", "genre_music")

CSV_FILE_PATH <-  "/Users/nickhilsethi/src/stat706-class-materials/final-project/terraform/project-infra/project_data.csv"
PROFIT_THRESHOLD <- 0.0
```

# Introduction
The goal of this paper is to investigate film profitability in Kaggle's ["The Movies Dataset"](https://www.kaggle.com/rounakbanik/the-movies-dataset), an aggregate dataset from GroupLens and TMDB containing information on revenue, budget, ratings, and various qualitative traits (e.g. genre, medium) for roughly ~30,000 films.

In particular, we hope to answer the following questions:

- How does genre affect profitability?
- How is critical reception related to profitability? 
- How is release date related to profitability? 
- How is budget related to profitability?

We begin from a complete set of predictors and use an iterative procedure to minimize the Akaike-Information Criterion (i.e. R's `step` function) in order to discover the most powerful predictors. We then examine the reduced model for significance, interpret the results, and construct prediction intervals for movies of varying characteristics.

# Data

## Transformations, Cleaning, and Schema
The Movies dataset contains three tables relevant to our analysis --  `movies_metadata`, `ratings`, and a join table `links`. The three tables have the following schemas:

```
MOVIES_METADATA {
  genres: [{ genreId: <int>,  name: <str> }],
  revenue: float,
  budget: float,
  imdbId: int
}

RATINGS {
  movieId: int,
  userId: int,
  rating: float,
}

LINKS {
  movieId: int,
  imdbId: int
}
```

These three tables were transformed into a single table `movies` used in the analysis for this paper. The schema of `movies` is as follows:
```
movies {
  imdb_id: int,
  budget: float,
  profit: float,
  release_date: date,
  average_rating: float,
  genre_action: boolean,
  genre_thriller: boolean,
  ...
  genre_music: boolean
}
```
Several transformations were performed to turn the three original tables into the usable format contained in `movies`. 

1. The genre column of `movies_metadata` is JSON, encoded as a list of pairs e.g. `[{genreId: 1, genreName: Action}, {genreId: 2, genreName: Comedy}]`, with each pair representing a genre the film is associated with; this column was converted to a set of boolean columns (e.g. `genre_action`) with `True` representing that the film belongs to that genre.

2. The `ratings` table contains movie ratings at the level of `(movieId, userId)` pairs, i.e. at the level of an individual critic's review; a grouping operation was performed to compute the average rating for each movie. 

3. A join was performed between `movies_metadata` and the grouped version of `ratings` using the join table `links`.

4. The `profit` column was added, defined as `revenue - budget` for each row; note the assumption here that all films use exactly there budget. 

5. Finally, rows with missing values are dropped. As can be seen from the chart below, the variable `is_complete` may be correlated with `profit`, (with incomplete rows typically having `0` profit); however, only a tiny fraction of rows are missing data, and this is not likely to influence the results in any major way.

```{r}
movies <- read.csv(file = CSV_FILE_PATH, header = TRUE, sep = ",")

# compute profits for each film
movies$profit <- movies$revenue - movies$budget

# cast the release_date field to a date type
movies$release_date <- as.Date(movies$release_date)

# cast genre columns to boolean
for (col in GENRE_COLUMNS) {
  movies[, col] <- lapply(movies[col], as.logical)
}

# now just restrict to the columns we need
features <- c(c("profit", "budget", "release_date", "average_rating"), GENRE_COLUMNS)

movies <- movies[, features]

movies$is_complete <- complete.cases(movies)

p1 <- ggplot(
  movies, aes(x=is_complete, y=profit)) + geom_jitter(width = .05, height=.05) + geom_point()
p2 <- ggplot(
  movies, aes(x=is_complete)) + geom_histogram(stat="count")
grid.arrange(p1, p2, nrow=1)

# now we drop the incomplete cases
movies <- movies[movies$is_complete == TRUE, ]
```

## Description of Data and Considerations for Analysis
The `movies` table contains four continuous variables -- `profit`, `budget`, `average_rating`, and `release_date`, as well as 21 boolean variables reperesenting genre classifications e.g. `genre_action`. The distributions of the four continuous variables, as well as scatter plots of the input variables against `profit` are shown below:

```{r}
p1 <- ggplot(movies, aes(x=profit)) + geom_histogram(bins=40)
p2 <- ggplot(movies, aes(x=average_rating)) + geom_histogram(bins = 25)
p3 <- ggplot(movies, aes(x=release_date)) + geom_histogram(bins = 25)
p4 <- ggplot(movies, aes(x=budget)) + geom_histogram(bins = 40)
p5 <- ggplot(movies, aes(x=budget, y=profit)) + geom_point()
p6 <- ggplot(movies, aes(x=release_date, y=profit)) + geom_point()
p7 <- ggplot(movies, aes(x=average_rating, y=profit)) + geom_point()
grid.arrange(p1, p2, p3, p4, p5, p6, p7, nrow=4)
```
We can glean a few interesting facts from inspection of the above graphs. Most of these variables appear to be normally distributed, with the exceptions of `profit` and `budget` which are heavily peaked at zero. The non-normality of these two variables becomes more visible if we restrict to films that are profit-positive, where we see the distributions more closely resembles an exponential than a gaussian:

```{r}
movies_profitable <- movies[movies$profit > 0, ]
p1 <- ggplot(movies_profitable, aes(x=profit)) + geom_histogram(bins=100)
p2 <- ggplot(movies_profitable, aes(x=budget)) + geom_histogram(bins=80)
grid.arrange(p1, p2, nrow=1)
```

The heavy skew of these variables towards zero can pose a problem for linear regression. The response variable being peaked at zero encourages the model to simply output zero; furthermore, the input `budget` variable being peaked at zero reduces the certainty of the models predictions for high budget films.  These concerns are actually confirmed in the `Discussion` section, see below.

A second issue results from the fact that `budget`, `average_rating`, and `release_date` are all correlated with profit, and are therefore correlated with each other; our procedure for model selection is minimizing AIC, but this may not yield the most interpretable model.

Finally, let's take a look at the genre columns, to get a sense of the relative frequencies:

```{r}

genre_counts <- read.csv('/Users/nickhilsethi/src/stat706-class-materials/final-project/terraform/project-infra/genre_counts.csv')
genre_counts <- genre_counts[genre_counts$genre %in% GENRE_COLUMNS, ]
ggplot(genre_counts, aes(x=genre, y=count)) + geom_bar(stat="identity") + theme(text = element_text(size=12), axis.text.x = element_text(angle=90, vjust=0.5, hjust = 1))
```

## Methods and Results

We begin by fitting a model with all available predictors; `budget`, `average_rating`, `release_date`, and every genre are included in the model. Then, we call `R`'s `step` function on the model, which iteratively drops variables by testing if removing them from the model lowers the `AIC`. 

Below is a summary of the results for the complete model:

```{r}
lmod <- lm(profit ~ ., data = movies)
sumary(lmod)
aic <- AIC(lmod)
copy <- sprintf("AIC for full model: %f", aic)
noquote(format(copy))
```


And see here for the summary results of the reduced model:

```{r}
lmod_reduced <- step(lmod, trace=0)
sumary(lmod_reduced)
aic <- AIC(lmod_reduced)
copy <- sprintf("AIC for reduced model: %f", aic)
noquote(format(copy))
```

The AIC minimzation process results in seven genre variables being dropped, yielding a decrease in AIC from $1166395$ to $1166388$. Both models have an $R^{2}$ of `.4`.

```{r}
qqnorm(rstandard(lmod_reduced))
qqline(rstandard(lmod_reduced))
```

# Discussion

## Model formulation
The model formulation and technique used here was chosen to "let the data speak"; that is, to determine the best predictors of `profit` without imposing too much speculation on model structure from the outset. For this reason, interaction effects between genres e.g. "action-comedy" or "musical-drama" were not accounted for. Doing so would entail arbitrary judgements about which interactions are valid "compound" genres; on the other hand, adding all pairwise interactions among 20 genres would add 190 additional predictors, with which comes numerous co-linearity problems and computational overhead.

## Interpretation

### Model Coefficients

Considerable information was gleaned from the data. Given a threshold of .05, all variables are significant in the final model with the exceptions of `release_date`, `fantasy`, `family`, and `foreign`.

The AIC minimization procedure eliminated 7 variables from the complete model, all of which are genres. Some general themes are observed in the set of variables which were removed; more niche genres such as `tv_movie` are removed for more mainstream ones such as `action`. Inspecting the output of the linear models above, we see the the predictors dropped by the `step` function typically had p-values well above .05 in the original model.

Turning to the signs of the coefficients, the only significant genres which have a positive effect on profit are `adventure`, `horror`, and `animation`. Genres `action`, `crime`, `drama`, `war`, and `thriller` have negative effects, perhaps indicating people tend to avoid films with stressful emotional content (though interestingly, `horror` has a positive coefficient). Genres `history` and `western` have negative coefficients as well, perhaps indicating an audience aversion to historical settings. The remaining variables `budget`, `release_date`, and `average_rating` all have positive effects on `profit.`. 

Regarding the magnitudes of the coefficients, it is critical to note the varying scales of the variables. Genre variables are binary (i.e. take value 0 or 1) and therefore have coefficients on the order of magnitude `10e6`, with the exception of `western`, which has an effect on the order of `10e7`. `budget` and `release_date` have effects on the order of `10e0` and `10e1`, respectively. It is worthy of note that the `average_rating` variable has one of the strongest effects on `profit`, with a coefficient of `4.0986e+06`; a single star increase in rating entails an increase in profit of $4,000,000. The magnitudes of the coefficients suggest that the best way to make a profitable film is to create an adventure 

### Diagnostics

The model errors are "heavy-tailed" rather than normally distributed. Though this has no pertinence to the sign and magnitude of our estimated coefficients, it does make our significance estimates and confidence intervals suspect. I attempted to solve this problem by restricting the analysis to films which were profit positive and log-transforming the `profit` variable, but this did not remediate this issue.

As we can see below, the residual is very much correlated with all the continuous variables after the regression. 

```{r}
movies$residual <- residuals(lmod_reduced)
movies$prediction <- predict(lmod_reduced)
p1 <- ggplot(data = movies, aes(x=residual)) + geom_histogram(bins=100)
p2 <- ggplot(data = movies, aes(x=profit, y=residual)) + geom_point()
p3 <- ggplot(data = movies, aes(x=budget, y=residual)) + geom_point()
p4 <- ggplot(data = movies, aes(x=average_rating, y=residual)) + geom_point()
grid.arrange(p1, p2, p3, p4,  nrow=3)
```

In fact the plots of above variables against the residuals closely resemble the plots of the variables against the response. This indicates the model is predicting zero for most films, likely being heavily driven by the vast majority of films which yield zero profit. As we can see below: 
```{r}
ggplot(data = movies, aes(x=prediction, y=profit)) + geom_point()
```

Let us briefly investigate a log transformation on the response. As we can see, dropping movies whose profit is not strictly greater than zero and then log transforming the profit column makes the data more closely resemble a gaussian:

```{r}
movies <- movies[movies$profit > PROFIT_THRESHOLD, ]
features <- c(c("profit", "budget", "release_date", "average_rating"), GENRE_COLUMNS)
movies <- movies[, features]

movies$log_profit <- log(movies$profit + 100)
ggplot(data = movies, aes(x=log_profit)) + geom_histogram()
```


Now let us run the same procedure as above, but with the response as `log(profit)`. Unfortunately this seems to create the opposite problem; the predictor is linear in `profit` but should be linear in `log(profit)`, and so we have the same interpretation problem as we had before.
```{r}
lmod <- lm(log_profit ~ ., data = movies)
lmod <- step(lmod, trace=0)
sumary(lmod)
copy <- sprintf("AIC for model with log transformation: %f", AIC(lmod))
noquote(format(copy))

movies$pred <- predict(lmod)
p1 <- ggplot(data = movies, aes(x=pred, y=log_profit)) + geom_point()
p2 <- ggplot(data = movies, aes(x=pred, y=profit)) + geom_point()

grid.arrange(p1, p2, nrow=1)
```

## Future Directions and Improvements

Changing the link function to account for heavy tails. 
