---
title: "Homework 1: R for Data Science / 3.2.4"
author: "Nickhil-Sethi"
date: "8/31/2020"
output:
  pdf_document: default
  html_document: default
---

Before completing the questions, let's load the `tidyverse` library.
```{r}
library(tidyverse)
```

## 1) Run `ggplot(data = mpg)`. What do you see?
Running the above command, we actually see a blank screen (see next page).
```{r}
ggplot(data = mpg)
```

## 2) How many rows are in `mpg`? How many columns?
`mpg` has 234 rows and 11 columns, as can be seen from printing the truncated dataframe:
```{r}
mpg
```


## 3) What does the `drv` variable describe? Read the help for `?mpg` to find out.

`drv` is a categorical variable describing type of drive train; values `f`, `r`, and `4` indicate front-wheel drive, rear-wheel drive, and four-wheel drive, respectively. 

## 4) Make a scatterplot `hwy` vs `cyl`.
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = cyl, y = hwy))
```

## 5) What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?
See next page for plot. It's difficult to interpret this plot because both `drv` and `class` are categorical variables. There is no order structure on either of the variables i.e. the way we embed them into the x and y axes is arbitrary, and so any 'trend' or 'line of fit' here is meaningless.
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = drv, y = class))
```
