---
title: "R for Data Science / 3.2.4"
author: "Nickhil-Sethi"
date: "8/31/2020"
output:
  pdf_document: default
  html_document: default
---

# R for Data Science, 3.2.4
```{r}
library(tidyverse)
```

## 1) Run `ggplot(data = mpg)`. What do you see?
Running the above command, we actually see a blank screen:
```{r}
ggplot(data = mpg)
```
# How many rows are in `mpg`? How many columns?
`mpg` has 234 rows and 11 columns, as can be seen from printing the truncated dataframe:
```{r}
mpg
```


# What does the `drv` variable describe? Read the help for `?mpg` to find out.

`drv` is a categorical variable describing type of drive train. From the help page of the `mpg` data set, `the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd`. 

# Make a scatterplot
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = cyl, y = hwy))
```

# Make a scatterplot of 
It's hard to make sense of this plot because both `drv` and `class` are categorical variables. There is no order structure on either of the variables i.e. the way we embed them into the x and y axes is arbitrary. 
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = drv, y = class))
```
